Improving Haskell Types with SMT

### 2015-03-13 docs/paper.tex
Add some examples.

The paper is organized as follows:  we start with a brief overview
of SMT solvers form a user's perspective (Section~\ref{smt}); this
section may be skipped by readers familiar with the tools.  Then,
in Section~\ref{GHC}, we introduce the basic concepts of GHC's constraints
solver, which is necessary for putting the rest of the paper in context.
Section~\ref{algorithm} contains the details of the algorithm for
integrating an SMT solver with GHC, and Section~\ref{other} explains
other theories that could be added to GHC using the same technique.
Finally, Section~\ref{modular-typechecker} discusses the possibility
of using the ideas in this paper, and previous work on the
implementation of SMT solver, to engineer a modular constraint solver.

>>>>

\subsection{Examples}

We illustrate the utility of the functionality provided by our
algorithm with a few short examples.  A very common example in
this area is to define a family of singleton types that links
type-level natural numbers to run-time constants that represent
them:
\begin{Verbatim}
data UNat :: Nat -> * where
  Zero :: UNat 0
  Succ :: UNat n -> UNat (n + 1)
\end{Verbatim}
Here we've used a unary representation of the natural numbers,
and each member of the family, \Verb"UNat n", has exactly one
inhabitant, namely the natural number \Verb"n" represented in
unary form.  Because we are using a GADT, we can pattern match
on the constructors of the type and gradually learn additional
information about the value being examined.

Next, we define a function to add two such numbers:
\begin{Verbatim}
uAdd :: UNat m -> UNat n -> UNat (m + n)
uAdd Zero x      = x
uAdd (Succ x) y  = Succ (uAdd x y)
\end{Verbatim}
While this is a simple definition, and we are unlikely
to have gotten it wrong, it is nice to know that GHC
is checking our work!  Had we made a mistake, for
example, by mis-typing the recursive call as \Verb"uAdd x x",
we would get a type error:
\begin{Verbatim}
  Could not deduce (((n1 + n1) + 1) ~ (m + n))
    from the context (m ~ (n1 + 1))
\end{Verbatim}
It is a small detail, but it is worth pointing out that
GHC would have been just as happy had we defined the type
of \Verb"Succ" like this:
\begin{Verbatim}
  Succ :: UNat n -> UNat (1 + n)
\end{Verbatim}
The difference is in the return type, in the one case it is
\Verb"n + 1", and in the other it is \Verb"1 + n".  While
to a human this looks like an insignificant difference,
in many systems it is significantly easier to work
with the one definition, but not the other.

The unary natural numbers are handy if we are planning to
do iteration.  However, if we are working on some sort of
divide-and-conquer algorithm, we often need to split the
input in two.  In this situation, a different family of
singletons is more useful:
\begin{Verbatim}
data BNat :: Nat -> * where
  Empty :: BNat 0
  Even  :: (1 <= n) => BNat n -> BNat (2 * n)
  Odd   :: BNat n -> BNat (2 * n + 1)
\end{Verbatim}
This is another singleton family, where the type \Verb"BNat n"
is inhabited by a single member, \Verb"n".  However, in this
case we learn different information by pattern examining
the values:  an \Verb"Empty" value may not be split,
an \Verb"Even" value may be split into two equal non-empty
parts, while an \Verb"Odd" value may be split into two parts,
and there will be one element left over.

So, how do we add such numbers?  There are more cases to
consider:
\begin{Verbatim}
bAdd :: BNat m -> BNat n -> BNat (m + n)
bAdd Empty x            = x
bAdd x Empty            = x
bAdd (Even x) (Even y)  = Even (bAdd x y)
bAdd (Even x) (Odd y)   = Odd  (bAdd x y)
bAdd (Odd x)  (Even y)  = Odd  (bAdd x y)
bAdd (Odd x)  (Odd  y)  = Even (bSucc (bAdd x y))
\end{Verbatim}
The correctness of this definition is much less obvious,
and we did make a couple of mistakes before getting it
right!  In the last case, we use the auxiliary function
\Verb"bSucc", which increments a binary natural number
by one:
\begin{Verbatim}
bSucc :: BNat m -> BNat (m + 1)
bSucc Empty     = Odd Empty
bSucc (Even x)  = Odd x
bSucc (Odd x)   = Even (bSucc x)
\end{Verbatim}

While these examples are somewhat simplistic, we hope
that they demonstrate the utility of the technology,
and show that using the extension feels quite natural
to a Haskell programmer.

\subsection{Structure of the Paper}

The rest of the paper is organized as follows:
we start with a brief overview
of SMT solvers form a user's perspective (Section~\ref{smt}); this
section may be skipped by readers familiar with the tools.  Then,
in Section~\ref{GHC}, we introduce the basic concepts of GHC's constraints
solver, which is necessary for putting the rest of the paper in context.
Section~\ref{algorithm} contains the details of the algorithm for
integrating an SMT solver with GHC, and Section~\ref{other} explains
other theories that could be added to GHC using the same technique.
Finally, Section~\ref{modular-typechecker} discusses the possibility
of using the ideas in this paper, and previous work on the
implementation of SMT solver, to engineer a modular constraint solver.
